---
title: "One month as a procrastinator on Stack Overflow"
author: "Florian Privé"
date: "July 27, 2017"
output:                    # DO NOT CHANGE
  prettydoc::html_pretty:  # DO NOT CHANGE
    theme: cayman          # DO NOT CHANGE
    highlight: github      # DO NOT CHANGE
---

Hello everyone, I'm 6103040 aka F. Privé.
In this post, I will give some insights about answering questions on Stack Overflow (SO) for a month.
One of the reason I've began frenetically answering questions on Stack Overflow was to procrastinate while finishing a scientific manuscript.

## My activity on Stack Overflow

We'll use David Robinson's package **stackr** to get data from Stack Overflow API.

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "80%", fig.asp = 0.75, fig.width = 8)
```

```{r}
# devtools::install_github("dgrtwo/stackr")
suppressMessages({
  library(stackr)
  library(tidyverse)
  library(lubridate)
})
```


### Evolution my SO reputation

```{r}
myID <- "6103040"

myRep <- stack_users(myID, "reputation-history", num_pages = 10)

(p <- myRep %>%
    arrange(creation_date) %>%
    ggplot(aes(creation_date, cumsum(reputation_change))) %>%
    bigstatsr:::MY_THEME() +
    geom_point() +
    labs(x = "Date", y = "Reputation", 
         title = "Evolution of my SO reputation over time"))

p +
  xlim(as.POSIXct(c(today() - months(1), today()))) +
  geom_smooth(method = "lm") +
  ggtitle("Evolution of my SO reputation over the last month")
```

### Analyzing my answers

```{r}
(myAnswers <- stack_users(myID, "answers", num_pages = 10,
                          fromdate = today() - months(1)) %>%
    select(-starts_with("owner")) %>%
    arrange(desc(score)) %>%
    as_tibble())
```

So it seems I've answered `r nrow(myAnswers)` questions over the past month. Interestingly, my answers with the greatest scores were not accepted. You can get a look at these using 

```{r, echo=FALSE, comment = ""}
browseAnswersText <- function(answers) {
  paste0("https://stackoverflow.com/questions/", answers$question_id) %>%
    paste0('"', ., '"') %>%
    paste(collapse = ",\n         ") %>%
    cat("sapply(c(", ., "), browseURL)", sep = "")
}

browseAnswersText(filter(myAnswers, score >= 5))
```

The first one is just translating some R code in Rcpp. The two other ones are **dplyr** questions.

```{r}
myAnswers %>% 
  group_by(score) %>%
  summarise(
    N = n(),
    acceptance_ratio = mean(is_accepted)
  )
```

My acceptance rate is quite bad. 


### Tags I'm involved in

```{r}
stack_users(myID, "tags", num_pages = 10) %>%
  select(name, count) %>%
  DT::datatable() 
```

<br/>

- I'm obviously answering only R questions

- Questions about data frames or **dplyr** are quite easy so that I've answered several of them

- I'm answering tags *r-bigmemory*, *rcpp*, *foreach*, *parallel-foreach* as I'm getting good at it because of the package I've developed (see [my previous post](https://privefl.github.io/blog/package-bigstatsr-statistics-with-matrices-on-disk-user-2017/)).


## Some insights from this experience

- **purrr** is badly received as a proxy of base R functions such as `sapply` and `lapply` (https://stackoverflow.com/questions/45101045/why-use-purrrmap-instead-of-lapply)

- People tend to use **dplyr** where base R functions are well-suited:

    - https://stackoverflow.com/questions/45244063/using-dplyr-to-replace-a-vector-of-names-with-new-names
    - https://stackoverflow.com/questions/45243363/dplyr-to-calculate-of-prevalence-of-a-variable-in-a-condition
    - https://stackoverflow.com/questions/44881723/replace-column-by-another-table
    - https://stackoverflow.com/questions/44995997/create-a-new-variable-using-dplyrmutate-and-pasting-two-existing-variables-for
    - https://stackoverflow.com/questions/45309455/mean-function-producing-same-result
    
- To avoid basic issues, I find very important to know base R classes and their accessors well (you just need to read [Advanced R](http://adv-r.had.co.nz/))

- The tidyverse solves lots of problems (you just need to read [R for Data Science](http://r4ds.had.co.nz/))

- Guiding to a solution is much more fun than just giving it (https://stackoverflow.com/questions/45308904). 


## Conclusion and bonuses

I think it has been a good experience to answer questions on SO for a month. 

I'm proud of [this algorithm written only with **dplyr**](https://stackoverflow.com/a/45302898/6103040) that automatically get you a combination of variables to form a [unique key](https://en.wikipedia.org/wiki/Unique_key) of a dataset. Also, I wanted to make a blog post about good practices for parallelization in R. I'm not sure how to do it and which format to use, but, for now, you can get [some good practices in one of my answer](https://stackoverflow.com/a/45196081/6103040). Finally, if you miss the previous infinite printing of a tibble, you can get a workaround [there](https://stackoverflow.com/a/44868837/6103040).
