---
title: "Making a team survey to get my colleagues hooked on R"
author: "Florian Privé"
date: "2016-10-07"
output:                    # DO NOT CHANGE
  prettydoc::html_pretty:  # DO NOT CHANGE
    theme: cayman          # DO NOT CHANGE
    highlight: github      # DO NOT CHANGE
---

In this post, I will talk about the presentation of R that I did today, in the first week of my PhD. 
Usually, it is a team-only presentation. Yet, other people came because they were interested in learning more about R.

## How I get this idea?

I get the idea of doing an R presentation while reading [Getting Your Colleagues Hooked on R](https://www.r-bloggers.com/getting-your-colleagues-hooked-on-r/) on [R-bloggers](https://www.r-bloggers.com/).
I began by following the 7 tips of this post to make my presentation, which was a good starting point.

After a while, I feared that a general presentation would not get my team interested in R. So, I decided to set up a [google form](https://goo.gl/forms/LREeX5NORBJlCrcC3) and ask them what they wanted to learn about R.
It was the way to get sure that they would care.

## Get results automatically

Because I was writing my R Markdown presentation while they were answering the google form, I decided that I should get (and show) the results automatically (only by re-knitting my presentation).

### To get the results

I used the `gsheet` package (one could also use the `googlesheets` package):
```{r, eval=FALSE}
library(pacman)
p_load(magrittr, longurl, gsheet)

responses <- "goo.gl/4zYmrw" %>% expand_urls %>% {gsheet2tbl(.$expanded_url)[, 2]}
```


### To get the different possible choices of the form

I got them directly from reading the website of the google form:
```{r, eval=FALSE}
p_load(gsubfn, stringr)

questions <- 
  "https://goo.gl/forms/LREeX5NORBJlCrcC3" %>%
  readLines(encoding = "UTF-8") %>%
  strapply(pattern = "\\[\"([^\"]*)\",,,,0\\]") %>%
  unlist
```
I couldn't get them directly from the googlesheet because google doesn't make the difference between a comma in the name of the choices and commas used to seperate multiple answers. If you know how to specify the separation when generating results from a google form, I'd like to know.

### To print the results directly in my presentation

I used the chunk option `results='asis'`:
```{r, eval=FALSE}
counts <- str_count(responses, coll(questions))
counts.lvl <- counts %>% unique %>% sort(decreasing = TRUE) %>% setdiff(0)

printf <- function(...) cat(sprintf(...))

for (n in counts.lvl) {
  if (n == 2) printf("\n***\n")
  printf("- for **%d** of you:\n", n)
  q.tmp <- questions[counts == n]
  for (q in q.tmp) {
    printf("    - %s\n", q)
  }
}
```
in order to generate markdown from R code.

### Getting the number of R packages on CRAN

I also wanted to show them how many package we have on CRAN, so I used:
```{r, eval=FALSE}
n <- readLines('https://cran.r-project.org/web/packages/') %>%
  gsubfn::strapply(
    paste("Currently, the CRAN package repository",
          "features ([0-9]+) available packages.")) %>%
  unlist
```
and printed `n` as inline R code.

## Conclusion

You can see the presentation [there](https://privefl.github.io/R-presentation/pResentation.html) and the corresponding Rmd file [there](https://privefl.github.io/R-presentation/pResentation.Rmd).

After finishing my presentation, I realized that most of what I presented, I learned it on R-bloggers. So, thanks everyone for the wonderful posts we get to read everyday!

If some of you think about other things that are important to know about R, I'd like to hear about them, just as personal curiosity.