---
title: "One year as a subscriber to Stack Overflow"
author: "Florian Privé"
date: "July 2, 2018"
output:                    # DO NOT CHANGE
  prettydoc::html_pretty:  # DO NOT CHANGE
    theme: cayman          # DO NOT CHANGE
    highlight: github      # DO NOT CHANGE
---

In this post, I follow up on a [previous post](https://privefl.github.io/blog/one-month-as-a-procrastinator-on-stack-overflow/) describing how last year in July, I spent one month mostly procrastinating on Stack Overflow (SO). We're already in July so it's time to get back to one year of activity on Stack Overflow. 

Am I still as much active as before? **What is my strategy for answering questions on SO?**

## My activity on Stack Overflow

Again, we'll use David Robinson's package {stackr} to get data from Stack Overflow API in R.

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "80%", fig.asp = 0.75, fig.width = 8)
```

```{r}
# devtools::install_github("dgrtwo/stackr")
suppressMessages({
  library(stackr)
  library(tidyverse)
  library(lubridate)
})
```


### Evolution of my SO reputation

```{r}
myID <- "6103040"

myRep <- stack_users(myID, "reputation-history", num_pages = 40,
                     fromdate = today() - years(1))

myRep %>%
  arrange(creation_date) %>%
  ggplot(aes(creation_date, cumsum(reputation_change))) +
  geom_point() +
  labs(x = "Date", y = "Reputation (squared transformed)",
       title = "Evolution of my SO reputation over the last year") + 
  bigstatsr::theme_bigstatsr()
```

So, it seems that my activity is slowing gently (my reputation is almost proportional to the square root of time). Yet, it is still increasing steadily; so what is my strategy for answering questions on SO?

### Tags I'm involved in

You'll have to wait for the answer to what is my strategy for answering questions on SO. For a hint, let's analyze the tags I'm involved in.

If we don't count my first month of activity:

```{r}
stack_users(myID, "tags", num_pages = 40,
            fromdate = today() - months(11)) %>%
  select(name, count) %>%
  as_tibble()
```

I'm obviously answering only R questions. The tags I'm mostly answering questions from are "performance", "rcpp", "parallel-processing", "foreach", "r-bigmemory" and "vectorization". 

### Performance

As you can see, all these tags are about performance of code.
I really enjoy performance problems (get the same result but much faster).

I can spend hours on a question about performance and am sometimes rewarded with a solution that is 2-3 order of magnitude faster (see e.g. [this other post](https://privefl.github.io/blog/performance-when-algorithmics-meets-mathematics/)).

I hope I could share my knowledge about performance through a tutorial in Toulouse next year.

## Conclusion and answer

So, the question was "What is my strategy for answering questions on SO?". And the answer is.. in the title: I am a subscriber.

I subscribe to tags on Stack Overflow. It has many benefits:

- you don't have to [rush to answer](https://meta.stackexchange.com/questions/9731/fastest-gun-in-the-west-problem) because questions you receive by mail are 30min-old (unanswered?) ones, so the probability that someone will answer at the same time as you is low.

- you can focus and what you're good at, what you're interested in, or just what you want to learn. For example, I subscribed to the very new tag "r-future" (for the R package {future}) because I'm interested in this package, even if I don't know how to use it yet. I had the chance to meet with its author, Henrik Bengtsson, at eRum2018 and he actually already knew me through parallel questions on SO :D.

However, some tags (like "performance" or "foreach") are relevant to many programming languages so that you would be flooded with irrelevant questions if subscribing directly to these tags. A simple solution to this problem is to subscribe to a feed of a combination of tags, like https://stackoverflow.com/feeds/tag?tagnames=r+and+foreach&sort=newest. I use [this website](https://blogtrottr.com/) to subscribe to feeds.

I will continue answering questions on SO, so see you there!

***

PS: I'm not sure you would get only unanswered questions with this technique.