---
title: "Using clustering to find points in an image"
author: "Florian Privé"
date: "November 20, 2018" # DO NOT USE Sys.Date()
output:                    # DO NOT CHANGE
  prettydoc::html_pretty:  # DO NOT CHANGE
    theme: cayman          # DO NOT CHANGE
    highlight: github      # DO NOT CHANGE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
par(mar = rep(0.1, 4))
```


In this post, I present my new package {img2coord}. This package can be used to retrieve coordinates from a scatter plot (as an image).    
Have you ever made a plot, saved it as a png and moved on? When you get back to it, it is sometimes difficult to read the values from this plot, especially if there is no grid inside the plot.    
This was a also a good way to practice with clustering.

## A very simple example

### Saving a plot as PNG

```{r}
file <- tempfile(fileext = ".png")
png(file, width = 600, height = 400)
set.seed(1)
plot(c(0, runif(20), 1))
dev.off()
```

### Reading the PNG in R

```{r}
img <- imager::grayscale(imager::load.image(file))
plot(img, rescale = FALSE, axes = FALSE)
```

### Get pixel indices from points

```{r}
library(Matrix)
img_mat <- Matrix(round(1 - as.matrix(img), 14), sparse = TRUE)
image(t(img_mat))
```

```{r}
list.contour <- img2coord:::get_contours(img_mat)
img_mat_in <- img2coord:::get_inside(img_mat, list.contour)
Matrix::image(t(img_mat_in))
```

```{r}
head(ind <- which(img_mat_in != 0, arr.ind = TRUE))
```

### Cluster indices of pixels

```{r}
set.seed(1)
km <- kmeans(ind, centers = 22)
plot(ind, pch = 20)
points(km$centers, pch = 20, col = "red")
```

Even when using the true number of clusters, *kmeans* get trapped in a local minimum (this is clearly not the best solution!), depending on the initialisation of centers. One possible solution would be to use many initialisations; let's try that.

```{r}
set.seed(1)
km <- kmeans(ind, centers = 22, nstart = 100, iter.max = 100)
plot(ind, pch = 20)
points(km$centers, pch = 20, col = "red")
```

It is better but not optimal.

### Using hclust to get centers

```{r}
get_centers <- function(points, clusters) {
  do.call("rbind", by(points, clusters, colMeans, simplify = FALSE))
}

hc <- hclust(d <- dist(ind))
centers <- get_centers(ind, cutree(hc, k = 22))
plot(ind, pch = 20)
points(centers, pch = 20, col = "red")
```

`hclust()` works well on this example.

### Get the number of clusters

What if we don't know the number of clusters (representing the initial points)? A statistic that could help us determine the number of clusters to use is the silhouette.

```{r}
K_seq <- seq(10, 30)
stat <- sapply(K_seq, function(k) {
  mean(cluster::silhouette(cutree(hc, k), d)[, 3])
})
plot(K_seq, stat, pch = 20); abline(v = 22, lty = 3)
```

## A less simple example


```{r}
file <- tempfile(fileext = ".png")
png(file, width = 600, height = 400)
set.seed(1)
y <- c(0, runif(100), 1)
plot(y, cex = runif(102, min = 0.5, max = 1.5))
dev.off()
```

```{r}
img <- imager::grayscale(imager::load.image(file))
plot(img, rescale = FALSE, axes = FALSE)
```

```{r}
img_mat <- Matrix(round(1 - as.matrix(img), 14), sparse = TRUE)
list.contour <- img2coord:::get_contours(img_mat)
img_mat_in <- img2coord:::get_inside(img_mat, list.contour)
ind <- which(img_mat_in != 0, arr.ind = TRUE)
```

```{r}
hc <- hclust(d <- dist(ind))
K_seq <- seq(50, 150)
stat <- sapply(K_seq, function(k) {
  mean(cluster::silhouette(cutree(hc, k), d)[, 3])
})
plot(K_seq, stat, pch = 20); abline(v = 102, lty = 3)
```

```{r}
(K_opt <- K_seq[which.max(stat)])
centers <- get_centers(ind, cutree(hc, k = K_opt))
plot(ind, pch = 20)
points(centers, pch = 20, col = "red")
```

The silhouette statistic is not working that well in this situation. With the true number of points, we would could get:

```{r}
centers <- get_centers(ind, cutree(hc, k = 102))
plot(ind, pch = 20)
points(centers, pch = 20, col = "red")
```

If someone has a better statistic to (automatically) find the number of clusters, please share and I'll update this post.

## Putting everything together as a package

After you get the center of all points (cluster of pixels), you can interpolate the values based on the values of axe ticks.

```{r}
coord <- img2coord::get_coord(
  file, 
  x_ticks = seq(0, 100, 20),
  y_ticks = seq(0, 1, 0.2),
  K_min = 50, K_max = 150
) 
```

This works better here because I combined the silhouette statistic with a gini coefficient (measure of dispersion) on the number of pixels in each cluster (assuming that they have approximately the same number). Let's have a look at the combined statistic:

```{r}
stat <- attr(coord, "stat")
plot(names(stat), stat, pch = 20); abline(v = 102, lty = 3) 
```

```{r}
round(coord$x, 2)
plot(coord$y, y, pch = 20); abline(0, 1, col = "red")
```

Not bad :-). If you don't get the right number of clusters the first time, you can use the plot generated by `img2coord::get_coord()` to adjust `K`.

Of course, this could be improved a lot. For example, this won't work for plots having a background color or some grid inside. Feel free to bring your ideas and propose some PR.

Have a look at [the GitHub repo](https://github.com/privefl/img2coord).
