---
title: "Fast R functions to get first principal components"
author: "Florian Privé"
date: "August 30, 2018" # DO NOT USE Sys.Date()
output:                    # DO NOT CHANGE
  prettydoc::html_pretty:  # DO NOT CHANGE
    theme: cayman          # DO NOT CHANGE
    highlight: github      # DO NOT CHANGE
---

In this post, I compare different approaches to get first principal components of large matrices in R.

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "70%", dev = "svg")
options(width = 95)
```

## Comparison

```{r, message=FALSE}
library(bigstatsr)
library(tidyverse)
```

### Data

```{r}
# Create two matrices, one with some structure, one without
n <- 20e3
seq_m <- c(1e3, 3e3, 10e3)
sizes <- seq_along(seq_m)
X <- E <- list()
for (i in sizes) {
  m <- seq_m[i]
  U <- matrix(0, n, 10); U[] <- rnorm(length(U))
  V <- matrix(0, m, 10); V[] <- rnorm(length(V))
  E[[i]] <- matrix(rnorm(n * m), n, m)
  X[[i]] <- tcrossprod(U, V) + E[[i]]
}
```

I use matrices of different sizes. Some are structured with 10 hidden components, and some with only random data.

### Optimized math library

I linked my R installation with OpenBLAS, an optimized parallel matrix library.

```{r}
(NCORES <- RhpcBLASctl::get_num_cores())
RhpcBLASctl::blas_set_num_threads(NCORES)
```

### Compared methods

```{r}
models <- tribble(
  ~method,                ~fun,                      ~params,
  "bigstatsr - 1 core",   bigstatsr::big_randomSVD,  list(k = 10),
  "bigstatsr - 6 cores",  bigstatsr::big_randomSVD,  list(k = 10, ncores = NCORES),
  "Rspectra",             RSpectra::svds,            list(k = 10),
  "irlba",                irlba::irlba,              list(nv = 10, nu = 10),
  "svd",                  svd::propack.svd,          list(neig = 10),
  "rsvd",                 rsvd::rsvd,                list(k = 10)
) %>%
  mutate(size = list(sizes), structured = list(c(TRUE, FALSE))) %>%
  unnest(size, .drop = FALSE) %>%
  unnest(structured, .drop = FALSE) %>%
  mutate(user_time = NA, real_time = NA, pcs = list(NA))
```

### Computing

```{r}
# Filling this data frame with times and PC scores for each method and dataset
for (i in rows_along(models)) {

  mat <- `if`(models$structured[[i]], X, E)[[models$size[[i]]]]

  time <- system.time({
    if (grepl("bigstatsr", models$method[[i]])) mat <- as_FBM(mat)
    res <- do.call(models$fun[[i]], args = c(list(mat), models$params[[i]]))
  })

  models[["user_time"]][[i]] <- time[1]
  models[["real_time"]][[i]] <- time[3]
  models[["pcs"]][[i]]  <- res
}

models <- mutate(models, size = seq_m[size])
```

### Timings

```{r, fig.asp=9/7, fig.width=7.5}
models %>%
  ggplot(aes(size / 1000, real_time, color = method)) +
  theme_bigstatsr() +
  geom_point(cex = 6) +
  geom_line(aes(linetype = method), lwd = 2) +
  facet_grid(structured ~ ., scales = "free") +
  theme(legend.position = c(0.25, 0.87),
        legend.key.width = unit(6, "line")) +
  labs(x = sprintf("ncol (x1000) (nrow = %d)", n), y = "Time (in seconds)",
       color = "Methods:", linetype = "Methods:")

models %>%
  filter(size == max(seq_m)) %>%
  select(method, structured, user_time, real_time)
```

### Errors

```{r}
true1 <- svd(X[[1]], nu = 10, nv = 10)
true2 <- svd(E[[1]], nu = 10, nv = 10)

bdiff <- function(x, y) {
  if (ncol(x) < ncol(y)) return(Inf)
  s = sign(x[1, ] / y[1, ])
  max(apply(sweep(x, 2, s, '*') - y, 2, crossprod))
}

models %>%
  filter(size == min(seq_m)) %>%
  mutate(error = map2_dbl(structured, pcs, ~{
    true <- `if`(.x, true1, true2)
    bdiff(.y$u, true$u)
  })) %>%
  select(method, structured, error)
```

## Conclusion

- Packages {rsvd} and {svd} don't give results precise enough when data is not structured.

- Packages {bigstatsr} and {irlba} are less precise (but precise enough!) than {RSpectra} because of a different tolerance parameter they use.

- Package {bigstatsr} is as fast as the other packages while not relying on matrix operations (see user timings above). So, even if you don't have your R installation linked to some optimized math library, you would get the same performance. On the contrary, the other methods would likely be much slower if not using such optimized library.

So, I highly recommend using package {RSpectra} to compute first principal components, because it is very fast and precise. Moreover, it works with e.g. sparse matrices. 
Yet, If you have very large matrices or no optimized math library, I would recommend to use my package {bigstatsr} that internally uses {RSpectra} but implements parallel matrix-vector multiplication in Rcpp for its data format **stored on disk**. To learn more on other features of R package {bigstatsr}, please have a look at [the package website](https://privefl.github.io/bigstatsr/).
