---
title: "Tip: Optimize your Rcpp loops"
author: "Florian Privé"
date: "December 29, 2016" # DO NOT USE Sys.Date()
output:                    # DO NOT CHANGE
  prettydoc::html_pretty:  # DO NOT CHANGE
    theme: cayman          # DO NOT CHANGE
    highlight: github      # DO NOT CHANGE
---

In this post, I will show you how to optimize your `Rcpp` loops so that they are 2 to 3 times faster than a standard implementation.

## Context

### Real data example

For this post, I will use a `big.matrix` which represents genotypes for 15,283 individuals, corresponding to the number of mutations (0, 1 or 2) at 287,155 different loci. Here, I will use only the first 10,000 loci (columns).

What you need to know about the `big.matrix` format:

- you can easily and quickly access matrice-like objects stored on disk,
- you can use different types of storage (I use type `char` to store each element on only 1 byte),
- it is column-major ordered as standard `R` matrices,
- you can access elements of a `big.matrix` using `X[i, j]` in `R`,
- you can access elements of a `big.matrix` using `X[j][i]` in `Rcpp`,
- you can get a `RcppEigen` or `RcppArmadillo` view of a `big.matrix` (see Appendix).
- for more details, go to [the GitHub repo](https://github.com/kaneplusplus/bigmemory).

Peek at the data:
```{r, include=FALSE}
options(width = 80)

require(bigsnpr)
celiac <- AttachBigSNP("../../bigsnpr/backingfiles/celiac_impute1_sub1.bk")
X <- sub.big.matrix(celiac$genotypes, lastCol = 1e4)
```

```{r}
print(dim(X))
print(X[1:10, 1:12])
```

### What I needed

I needed a fast matrix-vector multiplication between a `big.matrix` and a vector. Moreover, I could not use any `RcppEigen` or `RcppArmadillo` multiplication because I needed some options of efficiently subsetting columns or rows in my matrix (see Appendix). 

Writing this multiplication in `Rcpp` is no more than two loops:
```{Rcpp}
// [[Rcpp::depends(RcppEigen, bigmemory, BH)]]
#include <RcppEigen.h>
#include <bigmemory/MatrixAccessor.hpp>

using namespace Rcpp;

// [[Rcpp::export]]
NumericVector prod1(XPtr<BigMatrix> bMPtr, const NumericVector& x) {
  
  MatrixAccessor<char> macc(*bMPtr);

  int n = bMPtr->nrow();
  int m = bMPtr->ncol();

  NumericVector res(n);
  int i, j;
  
  for (j = 0; j < m; j++) {
    for (i = 0; i < n; i++) {
      res[i] += macc[j][i] * x[j];
    }
  }

  return res;
}
```

One test:
```{r}
y <- rnorm(ncol(X))

print(system.time(
  test <- prod1(X@address, y)
))
```

__What comes next should be transposable to other applications and other types of data.__

## Unrolling optimization

While searching for optimizing my multiplication, I came across [this Stack Overflow answer](http://stackoverflow.com/a/12289513/6103040).

Unrolling in action:
```{Rcpp}
// [[Rcpp::depends(RcppEigen, bigmemory, BH)]]
#include <RcppEigen.h>
#include <bigmemory/MatrixAccessor.hpp>

using namespace Rcpp;

// [[Rcpp::export]]
NumericVector prod4(XPtr<BigMatrix> bMPtr, const NumericVector& x) {
  
  MatrixAccessor<char> macc(*bMPtr);
  
  int n = bMPtr->nrow();
  int m = bMPtr->ncol();
  
  NumericVector res(n);
  int i, j;
  
  for (j = 0; j <= m - 4; j += 4) {
    for (i = 0; i < n; i++) { // unrolling optimization
      res[i] += (x[j] * macc[j][i] + x[j+1] * macc[j+1][i]) +
        (x[j+2] * macc[j+2][i] + x[j+3] * macc[j+3][i]);
    } // The parentheses are somehow important. Try without.
  }
  for (; j < m; j++) {
    for (i = 0; i < n; i++) {
      res[i] += x[j] * macc[j][i];
    }
  }
  
  return res;
}
```
```{r}
require(microbenchmark)

print(microbenchmark(
  PROD1 = test1 <- prod1(X@address, y),
  PROD4 = test2 <- prod4(X@address, y),
  times = 5
))

print(all.equal(test1, test2))
```

Nice! 
Let's try more. Why not using 8 or 16 rather than 4?

```{r, fig.align='center'}
Rcpp::sourceCpp('../code/prods.cpp')

print(bench <- microbenchmark(
  PROD1 = prod1(X@address, y),
  PROD2 = prod2(X@address, y),
  PROD4 = prod4(X@address, y),
  PROD8 = prod8(X@address, y),
  PROD16 = prod16(X@address, y),
  times = 5
))

time <- summary(bench)[, "median"]
step <- 2^(0:4)
plot(step, time, type = "b", xaxt = "n", yaxt = "n", 
     xlab = "size of each step")
axis(side = 1, at = step)
axis(side = 2, at = round(time))
```


## Conclusion

We have seen that unrolling can dramatically improve performances on loops.
Steps of size 8 or 16 are of relatively little extra gain compared to 2 or 4. 

As pointed out in the SO answer, it can behave rather differently between systems. So, if it is for your personal use, use the maximum gain (try 32!), but as I want my function to be used by others in a package, I think it's safer to choose a step of 4.

## Appendix

You can do a `big.matrix`-vector multiplication easily with `RcppEigen` or `RcppArmadillo` (see [this code](../prods2.cpp)) but it lacks of efficient subsetting option.

Indeed, you still can't use subsetting in `Eigen`, but this will come as
said in [this feature request](http://eigen.tuxfamily.org/bz/show_bug.cgi?id=329). For `Armadillo`, you can but it is rather slow:

```{r}
Rcpp::sourceCpp('../code/prods2.cpp')

n <- nrow(X)
ind <- sort(sample(n, size = n/2))

print(microbenchmark(
  EIGEN = test3 <- prodEigen(X@address, y),
  ARMA = test4 <- prodArma(X@address, y),
  ARMA_SUB = test5 <- prodArmaSub(X@address, y, ind - 1),
  times = 5
))

print(all(
  all.equal(test3, test), 
  all.equal(as.numeric(test4), test),
  all.equal(as.numeric(test5), test[ind])
))
```


