---
title: "Using clustering to find points in an image"
author: "Florian Privé"
date: "November 27, 2018" # DO NOT USE Sys.Date()
output:                    # DO NOT CHANGE
  prettydoc::html_pretty:  # DO NOT CHANGE
    theme: cayman          # DO NOT CHANGE
    highlight: github      # DO NOT CHANGE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", out.width = "80%")
```


In this post, I present my new [package {img2coord}](https://github.com/privefl/img2coord). This package can be used to retrieve coordinates from a scatter plot (as an image).

```{r, eval=FALSE}
devtools::install_github("privefl/img2coord")
```


Have you ever made a plot, saved it as a png and moved on? When you come back to it, it is sometimes difficult to read the values from this plot, especially if there is no grid inside the plot.    
Making this package was also a good way to practice with clustering.

## A very simple example

### Saving a plot as PNG

```{r}
file <- tempfile(fileext = ".png")
png(file, width = 600, height = 400)
set.seed(1)
plot(c(0, runif(20), 1))
dev.off()
```

### Reading the PNG in R

```{r}
(img <- magick::image_read(file))
```

### Get pixel indices from points

```{r}
## grayscale
img_mat <- img2coord:::img2mat(img)
dim(img_mat)
```

```{r}
list.contour <- img2coord:::get_contours(img_mat)
img_mat_in <- img2coord:::get_inside(img_mat, list.contour)
dim(img_mat_in)
```

```{r}
head(ind <- which(img_mat_in > 0, arr.ind = TRUE))
```

### Cluster pixel indices

```{r}
set.seed(1)
km <- kmeans(ind, centers = 22)
```

```{r}
library(ggplot2)
myplot <- function(points, centers) {
  p <- ggplot() + 
    geom_tile(aes(col, row), data = as.data.frame(points)) + 
    geom_point(aes(col, row), data = as.data.frame(centers), col = "red") + 
    bigstatsr::theme_bigstatsr() + 
    coord_equal()
  print(p)
}
myplot(ind, km$centers)
```

Even when using the true number of clusters, *kmeans* get trapped in a local minimum (this is clearly not the best solution!), depending on the initialisation of centers. One possible solution would be to use many initialisations; let's try that.

```{r}
set.seed(1)
km <- kmeans(ind, centers = 22, nstart = 100, iter.max = 100)
myplot(ind, km$centers)
```

It is better but not optimal.

### Using hclust to get centers

```{r}
get_centers <- function(points, clusters) {
  do.call("rbind", by(points, clusters, colMeans, simplify = FALSE))
}

d <- dist(ind)
hc <- hclust(d)
centers <- get_centers(ind, cutree(hc, k = 22))
myplot(ind, centers)
```

`hclust()` works well for this example.

### Get the number of clusters

What if we don't know the number of clusters (representing the initial points)? A statistic that could help us determine the number of clusters to use is the silhouette.

```{r}
K_seq <- seq(10, 30)
stat <- sapply(K_seq, function(k) {
  mean(cluster::silhouette(cutree(hc, k), d)[, 3])
})
plot(K_seq, stat, pch = 20); abline(v = 22, lty = 3)
```

## A less simple example


```{r}
file <- tempfile(fileext = ".png")
png(file, width = 600, height = 400)
set.seed(1)
y <- c(0, runif(100), 1)
plot(y, cex = runif(102, min = 0.5, max = 1.5))
dev.off()
```

```{r}
(img <- magick::image_read(file))
```

```{r}
img_mat <- img2coord:::img2mat(img)
list.contour <- img2coord:::get_contours(img_mat)
img_mat_in <- img2coord:::get_inside(img_mat, list.contour)
ind <- which(img_mat_in > 0, arr.ind = TRUE)
```

```{r}
hc <- flashClust::hclust(d <- dist(ind))
K_seq <- seq(50, 150)
stat <- sapply(K_seq, function(k) {
  mean(cluster::silhouette(cutree(hc, k), d)[, 3])
})
plot(K_seq, stat, pch = 20); abline(v = 102, lty = 3)
```

```{r}
(K_opt <- K_seq[which.max(stat)])
centers <- get_centers(ind, cutree(hc, k = K_opt))
myplot(ind, centers)
```

The silhouette statistic is giving a good yet not optimal solution in this situation. Using the true number of points, we would get:

```{r}
centers <- get_centers(ind, cutree(hc, k = 102))
myplot(ind, centers)
```

If someone has a better statistic to (automatically) find the number of clusters, please share it and I'll update this post.

## Putting everything together as a package

Finally, after you get the center of all points (pixel clusters), you can interpolate the values based on the values of axe ticks.

```{r, fig.width=9}
coord <- img2coord::get_coord(
  file, 
  x_ticks = seq(0, 100, 20),
  y_ticks = seq(0, 1, 0.2),
  K_min = 50, K_max = 150
) 
```

This works better here because I combined the silhouette statistic with a gini coefficient (measure of dispersion) of the number of pixels in each cluster (assuming that they should have approximately the same number). Let's have a look at the combined statistic:

```{r}
stat <- attr(coord, "stat")
plot(names(stat), stat, pch = 20); abline(v = 102, lty = 3) 
```

If you don't get the right number of clusters the first time, you can use the plot generated by `img2coord::get_coord()` to adjust `K`.

```{r, fig.width=9}
coord <- img2coord::get_coord(
  file, 
  x_ticks = seq(0, 100, 20),
  y_ticks = seq(0, 1, 0.2),
  K = 102  ## 99 + 3
) 
```

Let's verify the coordinates we get:

```{r}
round(coord$x, 2)
plot(coord$y, y, pch = 20); abline(0, 1, col = "red")
```

## Handling large images

```{r}
url <- "https://goo.gl/K6Y7D1"
library(img2coord)
(img <- img_read(url))
```

```{r, error=TRUE}
get_coord(img, seq(0, 20, 5), seq(94, 102, 2), K_min = 40, K_max = 80)
```

The green points are spanning 21,358 pixels, which could be a lot to process, depending on your computer. To solve this problem, you can do:

```{r, fig.width=9}
img %>%
 img_scale(0.4) %>%
 get_coord(seq(0, 20, 5), seq(94, 102, 2), K_min = 40, K_max = 80)
```


## Conclusion

We have seen that `hclust()` was performing better than `kmeans()` (for this example). For some reason I don't understand yet, initializing `kmeans()` with centers from `hclust()` works even better. 

Then, we have seen how to determine the number of clusters. Finally, we have seen that using a particular statistic, specifically designed for this problem, improved the solution.

Of course, this could be improved a lot. For example, this won't work for plots having a background color or some grid inside. Feel free to bring your ideas. BTW, thanks [Robin](https://twitter.com/robincura) who brought some nice ideas that improved this package a lot.

Have a look at [the GitHub repo](https://github.com/privefl/img2coord).
