---
title: "On the ifelse function"
author: "Florian Privé"
date: "October 15, 2016" # DO NOT USE Sys.Date()
output:                    # DO NOT CHANGE
  prettydoc::html_pretty:  # DO NOT CHANGE
    theme: cayman          # DO NOT CHANGE
    highlight: github      # DO NOT CHANGE
---

In this post, I will talk about the [**ifelse** function](https://stat.ethz.ch/R-manual/R-devel/library/base/html/ifelse.html), which behaviour can be easily misunderstood, as pointed out in [my latest question on SO](http://stackoverflow.com/questions/40026975/subsetting-with-negative-indices-best-practices). I will try to show how it can be used, and misued. We will also check if it is as fast as we could expect from a vectorized base function of R.

```{r setup, include=FALSE}
options(width = 90)
```


## How can it be used?

The first example comes directly from the R documentation:

```{r}
x <- c(6:-4)
sqrt(x)  #- gives warning
sqrt(ifelse(x >= 0, x, NA))  # no warning
```
So, it can be used, for instance, to handle special cases, in a vectorized, succinct way.

The second example comes from the [vignette of Rcpp Sugar](https://cran.r-project.org/web/packages/Rcpp/vignettes/Rcpp-sugar.pdf):

```{r}
foo <- function(x, y) {
  ifelse(x < y, x*x, -(y*y))
}
foo(1:5, 5:1)
```

So, it can be used to construct a vector, by doing an element-wise comparison of two vectors, and specifying a custom output for each comparison.

A last example, just for the pleasure:
```{r}
(a <- matrix(1:9, 3, 3))
ifelse(a %% 2 == 0, a, 0)
```

## How can it be misused?

I think many people think they can use `ifelse` as a shorter way of writing an `if-then-else` statement (this is a mistake I made). For example, I use:
```{r, eval=FALSE}
legend.pos <- ifelse(is.top, ifelse(is.right, "topright", "topleft"),
                     ifelse(is.right, "bottomright", "bottomleft"))
```
instead of:
```{r, eval=FALSE}
if (is.top) {
  if (is.right) {
    legend.pos <- "topright"
  } else {
    legend.pos <- "topleft"
  }
} else {
  if (is.right) {
    legend.pos <- "bottomright"
  } else {
    legend.pos <- "bottomleft"
  }
}
```

That works, but this doesn't:
```{r}
ifelse(FALSE, 0, 1:5)
```

Indeed, if you read carefully the R documentation, you see that `ifelse` is returning a vector of the same length and attributes as the condition (here, of length 1).

If you really want to use a more succinct notation, you could use 
```{r}
`if`(FALSE, 0, 1:5)
```

If you're not familiar with this notation, I suggest you read [the chapter about functions in book _Advanced R_](http://adv-r.had.co.nz/Functions.html).

## Benchmarks

### Reimplementing 'abs'

```{r}
abs2 <- function(x) {
  ifelse(x < 0, -x, x)
}
abs2(-5:5)
```

```{r}
library(microbenchmark)
x <- rnorm(1e4)

print(microbenchmark(
  abs(x), 
  abs2(x)
))
```

### Comparing with C++

Consider the Rcpp Sugar example again, 4 means to compute it:

```{r engine='Rcpp'}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector fooRcpp(const NumericVector& x, const NumericVector& y) {
  int n = x.size();
  NumericVector res(n);
  double x_, y_;
  for (int i = 0; i < n; i++) { 
    x_ = x[i];
    y_ = y[i];
    if (x_ < y_) {
      res[i] = x_*x_;
    } else {
      res[i] = -(y_*y_);
    }
  }
  return res;
}
```

```{r}
fooRcpp(1:5, 5:1)
```

```{r engine='Rcpp'}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export]]
NumericVector fooRcppSugar(const NumericVector& x, 
                           const NumericVector& y) {
  return ifelse(x < y, x*x, -(y*y));
}
```

```{r}
fooRcppSugar(1:5, 5:1)
```

```{r}
foo2 <- function(x, y) {
  cond <- (x < y)
  cond * x^2 - (1 - cond) * y^2
}
foo2(1:5, 5:1)
```


```{r}
x <- rnorm(1e4)
y <- rnorm(1e4)
print(microbenchmark(
  foo(x, y),
  foo2(x, y),
  fooRcpp(x, y),
  fooRcppSugar(x, y)
))
```

Even if it is a vectorized base R function, `ifelse` is known to be slow.

## Conclusion

Beware when you use the `ifelse` function. Moreover, if you make a substantial number of calls to it, be aware that it isn't very fast, but it exists at least 3 faster alternatives to it.